---
title: "Research Ethics Review Committee Anual Report 2021-2022"
date: "`r Sys.Date()`"
author: "Koen Leuveld and René Bekkers"
output: bookdown::pdf_document2
header-includes:
   - \usepackage{booktabs}
---


```{r, include=FALSE}



# load helper functions
source("rerc_selfcheck_overview_helpers.r")
source("rerc_selfcheck_datacleaning.R")


# source("rerc_selfcheck_qualtricsdownload.R") 


```




# Introduction
This memo reports research ethics review activities at the Faculty of Social Sciences at Vrije Universiteit Amsterdam in the academic year 2021/2022. Since 2016 the RERC has reviewed ethics issues of research by faculty members. The report gives an overview of the composition of the RERC and its activities, including the number self-checks done by faculty members and students and of the applications for full review. 

<!-- Rene: RB attended NETHICS meetings. 
A special meeting of the BMDO/OPCO, with action plan
No letters to the FB -->

# Membership Composition
The RERC consists of five members, one from each department of the Faculty. The RERC is supported by a secretary supplied by the Research Support Office: Koen Leuveld, who is also Data Steward and Privacy Champion. The members, and the dates by which their appointments end, are listed in Table 1. Throughout the academic year (September - June) members meet on a monthly basis to discuss ethics issues and policies within the faculty.

```{r,  echo=FALSE, message = FALSE}
read_csv("data/admin/Term_limits.csv") %>%
kable(booktabs = T, 
      longtable=F,
      caption="Composition of the RERC",
      position="h") %>%
kable_styling(latex_options = "striped") %>%
add_footnote(
  label = "Peter Versteeg will be replaced by Luisa Schneider, effective 1.9.2022",
  notation = "none"
)
```

```{r sc-staff-by-year, include=FALSE}

get_selfchecks <- function(var, position, value, outcome=NULL) {
   data <- self_check_data 

   if(!is.null(outcome)){
      data <- 
         data %>%
         filter(Outcome == outcome)
   }

   data %>%
   filter(Position == position) %>%
   filter({{var}} == value) %>%
   summarize(n = n())  %>%
   pull(1)

}

get_fullreviews <- function(var, value) {
   data <- full_review_data 

   data %>%
   filter({{var}} == value) %>%
   summarize(n = n())  %>%
   pull(1)

}

get_fullreviews(Academic_Year,"2022-2023")

```


# Research

A priority of the RERC has been to make the ethics review procedure more 
accessible. 
This starts with the self-check, which should not take much time to fill in and
should not cause confusion. 
To achieve this, the RERC has changed the self-check late in the 2021-2022 
Academic Year (28-03-2022).
Figure \@ref(fig:sc-staff-plot) and Table \@ref(tab:sc-staff-tab) show that 
the number of self-checks has increased somewhat
since then: from `r get_selfchecks(Academic_Year,"Staff","2021-2022")` in 
2021-2022 to `r get_selfchecks(Academic_Year,"Staff","2022-2023")` in 2022-2023. 
Of course this increase may have been caused by other reasons that the work of 
the RERC to make the procedures more accessible.
Possible alternative explanations include growth in staff or number or an
increase in research projects. 
The tables in the appendix explore this in more detail. 
In 2022 (the last year for which data is available in the SEP tables) 
there were more self-checks per research FTE (Table \@ref(tab:sc-staff-fte)) 
and per publication (Table \@ref(tab:sc-staff-outputs)).

Whatever the cause of the increased number of self-checks, they are good news. 
As it stands now, the self-check is one of the few monitoring tools available 
to the faculty to see how many researchers are using sensitive data or data from
vulnerable subject pools. 
More researchers taking the self-check thus means that the faculty has a better
view on what's happening.
In order to both increase the quality of monitoring, and the accessiblity of the
review procedure, the RERC will take part in a pilot by the library to build a
new system to replace the self-check. This is expected to start in the fall of 
2023.

```{r sc-staff-plot, fig.cap = "Number of completed self-checks by staff", echo=FALSE, message = FALSE}

self_check_data %>%
   filter(Position == "Staff") %>%
   collapse_df(Quarter) %>%
   filter(Quarter != "2023-3") %>%
   crop_df(Quarter,16) %>%
   ggplot(aes(x=Quarter, y=N,group=1)) +
      geom_line() +          
      geom_smooth(method="loess",formula = y ~ x,se=FALSE) +           
      theme(axis.text.x = element_text(angle = 90)) #+
      #ggtitle("Number of completed self-checks by staff") 
```

```{r sc-staff-tab, echo=FALSE, message = FALSE}

self_check_data %>%
   filter(Position == "Staff") %>%
   crop_df(Academic_Year,4) %>%
   tabyl(Department,Academic_Year)%>%
   adorn_totals(where=c("row","col")) %>%
   standard_kable(caption="Number of completed checks by staff and PhD candidates in the last 14 months") %>%
   row_spec(5,hline_after=TRUE)

```
Out of these `r get_selfchecks(Academic_Year,"Staff","2022-2023")`
self-checks, there were 
`r get_selfchecks(Academic_Year,"Staff","2022-2023","Review needed")` 
indications for full review (for an overview of the issues present, 
see Table (\@ref(tab:sc-issues)).


Not all self-checks indicating the need for full review have resulted in 
an application. 
The RERC only received `r get_fullreviews(Academic_Year,"2022-2023")` 
applications.
This is by design: one of the intended purposes of the self-check is to allow
researchers to change their research design to precent ethics issues altogether.








Out of these N, P proceeded to request full review.


# Education

# Appendix


```{r sc-staff-fte, echo=FALSE, message = FALSE}

# Staff with FTE
fte <- 
   sep %>%
   select(Department,Year,FTE) %>%
   crop_df(Year,3)


self_check_data %>%
   filter(Position == "Staff") %>%
   tabyl(Year,Department) %>%
   pivot_longer(cols = !Year,names_to="Department", values_to = "N") %>%
   inner_join(fte) %>%
   compute_totals(Year) %>%
   mutate(prop = N/FTE * 100) %>%
   pivot_wider(names_from = Year, values_from = c(N,FTE,prop))%>%
   order_columns_by_year %>%
   make_double_header_table(caption = "Number of completed checks by staff (N) 
                                       compared to research FTE, by department",
                            colnames = c("N","FTE","%"))

```



```{r sc-staff-outputs, echo=FALSE, message = FALSE}

# Staff with outputs

outputs <-
   sep %>%
   select(Department,Year,Publications) %>%
   crop_df(Year,3)



self_check_data %>%
   filter(Position == "Staff") %>%
   tabyl(Year,Department) %>%
   pivot_longer(cols = !Year,names_to="Department", values_to = "N") %>%
   inner_join(outputs) %>%
   compute_totals(Year) %>%
   mutate(prop = N/Publications * 100) %>%
   pivot_wider(names_from = Year, values_from = c(N,Publications,prop))%>%
   order_columns_by_year %>%
   make_double_header_table(caption = "Number of completed checks by staff (N) 
                                       compared to publications, by department",
                            colnames = c("N","Pub.","%"))


```

```{r sc-issues, echo=FALSE, message = FALSE}


self_check_data %>%
   filter(Academic_Year == "2022-2023") %>%
   filter(Position == "Staff") %>%
   filter(Outcome == "Review needed") %>%
   select(Name, Project,all_of(ethics_issues$new),Outcome) %>%
   pivot_longer(cols = consent:confidentiality, names_to="item", values_to = "N") %>%
   group_by(item) %>%
   summarize(n = sum(N)) %>%
   rename(new=item) %>%
   right_join(ethics_issues) %>%
   select(Issue,n) %>%
   standard_kable(caption="Ethics issues present in research 2022-2023") %>%
   add_footnote(
     label = "The total number of issues is larger than the number of self-checks requiring full review, because some projects have more than issue",
     notation = "none")

```